Quantifying the impacts of bias in landcover data on global change analyses

### 2015-08-27 paper/main.tex
Into results in manuscript. Updated cropland bias maps, bias statistics plots. Added Rmd for miscellaneous supporting calculations.

%\usepackage{pnastwoF}
\usepackage{cite}
\usepackage{lineno}
\setlength\linenumbersep{4pt}

@@

We compared our reference percent cropland estimates to those created from four satellite-derived landcover datasets. We obtained South Africa's 30 m resolution National Landcover map (SA-LC) for 2009 \cite{sanbi_national_2009}, the 500 m resolution MODIS Landcover for 2011 \cite{land_processes_distributed_active_archive_center_lp_daac_modis_2011, friedl_modis_2010}, the 300 m resolution GlobCover 2009 \cite{arino_global_2012}, and the new 1 km Geo-wiki hybrid-fusion cropland map for Africa \cite{fritz_mapping_2015}. We chose these particular datasets because they are nearly contemporaneous with our reference data, and represent the major types of landcover products used by researchers: SA-LC typifies the higher resolution, Landsat-derived maps that are developed individually for many countries \cite{fry_completion_2009},  MODIS and GlobCover are widely used global-scale products \cite{gross_monitoring_2013,shackelford_conservation_2015}, while Geo-Wiki incorporates the first three datasets and is the current state of the art for agricultural landcover maps. We extracted the cropland classes from the first three datasets and converted these to 1 km resolution percent cropland estimates (hereafter simply ``cropland maps''), resulting in 4 maps to compare to the reference.

\vspace{-0.5 cm}
\section{Quantifying Error}
We used these maps to first quantify error in cropland area estimates. We calculated error as the difference between the reference and cropland map at different scales of aggregation (1 to 100 km), to determine the extent of bias in each map and how it varies with scale. Next, we assessed how map error relates to cover patterns in agricultural  landscapes, by examining how map error over cropped pixels correlates with the actual amount of cropland.

We undertook five further analyses to investigate how map error can impact assessments that are founded on landcover maps. These include first-order analyses, in which values for a variable of interest are mapped to  particular cover type(s), and second-order analyses, in which a process model draws on the cover types' values to calculate an output value. We created four datasets to represent second order analyses. The first was a series of maps of vegetated carbon stocks created following the methodology of Ruesch and Gibbs' \cite{ruesch_new_2008}. The second was cropland percentage maps, which, following Ramankutty et al \cite{ramankutty_farming_2008} were adjusted so that their total areas matched the reported provincial-level cropland area. Using these adjusted cropland percentage maps, we followed the methods of Monfreda et al \cite{monfreda_farming_2008} to disaggregate census-reported maize harvested area and yields. We then compared differences between total carbon stock estimates calculated from the reference map with those from the four cropland maps, and again examined how these differences changed as a function of aggregation scale.  We made the same comparisons for total maize harvested area, average yield, and total production.

For the second-order analyses, we examined how cropland cover errors influence 25 km resolution monthly evapotranspiration estimates produced using the Variable Infiltration Capacity \cite{liang_simple_1994}. For this example, we used the cropland maps to adjust the seasonally varying, landcover-specific leaf area index (LAI) values that VIC uses to partition water vapor fluxes into their evaporative and transpirative components. In the second example, we examined how these errors can impact the parameterization of a computationally intensive, spatially explicit agent-based model of food security. In this case, we used the cropland maps to allocate farmland to model agents representing individual households in political districts, with each agent's initial holdings assigned as a function of the total cropland available in the district, and the proximity of cropland within a specified distance of agent's location.

%\vspace{-0.5 cm}
\section{Results}

\subsection{Percent cropland estimates}

>>>>

%\usepackage{pnastwoF}
%\usepackage{cite}
\usepackage[numbers,round,compress]{natbib}
\bibpunct{(}{)}{,}{n}{,}{,}  % https://xianblog.wordpress.com/tag/natbib/ (allows natbib with PNAS)
\usepackage{lineno}
\setlength\linenumbersep{2pt}

@@

We compared our reference percent cropland estimates to those created from four satellite-derived landcover datasets. We obtained South Africa's 30 m resolution National Landcover map (SA-LC) for 2009 \cite{sanbi_national_2009}, the 500 m resolution MODIS Landcover for 2011 \cite{land_processes_distributed_active_archive_center_lp_daac_modis_2011, friedl_modis_2010}, the 300 m resolution GlobCover 2009 \cite{arino_global_2012}, and the new 1 km Geo-wiki hybrid-fusion cropland map for Africa \cite{fritz_mapping_2015}. We chose these particular datasets because they are nearly contemporaneous with our reference data, and represent the major types of landcover products used by researchers: SA-LC typifies the higher resolution, Landsat-derived maps that are developed individually for many countries \cite{fry_completion_2009},  MODIS and GlobCover are widely used global-scale products \cite{gross_monitoring_2013,shackelford_conservation_2015}, while Geo-Wiki incorporates the first three datasets and is the current state of the art for agricultural landcover maps. We extracted the cropland classes from the first three datasets and converted these to 1 km resolution percent cropland estimates (hereafter simply the ``landcover maps''), resulting in 4 maps to compare to our reference cropland map (the ``reference map'').

\vspace{-0.5 cm}
\section{Quantifying Error}
We used these maps to first quantify error in cropland area estimates. We calculated error as the difference between the reference and landcover maps at different scales of aggregation (1 to 100 km), in order to estimate bias and how it varies with scale. Next, we assessed how bias correlates with the amount of cropland cover in agricultural landscapes, to gain insight into how landscape patterns may affect error.

We undertook five further analyses to investigate how map error can impact assessments founded on landcover maps. These include first-order analyses, in which values for a variable of interest are mapped to particular cover type(s), and second-order analyses, in which a process model draws on the cover types' values to calculate an output value. We created four datasets to represent second order analyses. The first was a series of maps of vegetated carbon stocks created following the methodology of Ruesch and Gibbs' \cite{ruesch_new_2008}. The second was constrained cropland percentage maps, which, following Ramankutty et al \cite{ramankutty_farming_2008} were adjusted so that their total cropland areas matched provincial-level reported cropland totals. Using these adjusted cropland percentage maps, we disaggregated district-reported maize harvested area and yields \cite[following][]{monfreda_farming_2008} . We then compared differences between total carbon stock estimates calculated from the reference map with those from the four cropland maps, and again examined how these differences changed as a function of aggregation scale.  We made the same comparisons for total maize harvested area, average yield, and total production.

For the second-order analyses, we examined how cropland cover errors influence 25 km resolution monthly evapotranspiration estimates produced using the Variable Infiltration Capacity \cite{liang_simple_1994} land surface hydrology model. For this example, we used the cropland maps to adjust the seasonally varying, landcover-specific leaf area index (LAI) values that VIC uses to partition water vapor fluxes into their evaporative and transpirative components. In the second example, we examined how these errors can impact the parameterization of an agent-based food security model \cite{chen_dependency_2013}. Spatially-explicit, agent-based models are frequently employed in land change science, and require an initialization step to assign landscape resources to model agents  \cite[e.g.][]{manson_agent-based_2007,evans_multi-scale_2004,kelley_relative_2011}. In this case, we used the cropland maps to allocate farmland to agents representing individual households in political districts, with the model assigning each household its initial cropland holdings using a function that considers total district cropland and how much cropland is near the agent's location.

\vspace{-0.5 cm}
\section{Results}
\subsection{Bias and its correlates}

We created the 1 km reference and landcover maps, removing all croplands marked as communal or smallholder farmland in the reference vector maps (individual fields were not mapped with the same precision) or as plantation forestry (SI), and then aggregated each map to 5, 10, 25, 50, and 100 km resolutions. We then subtracted each landcover map from the reference map at each scale of aggregation (Fig. 1).

\vspace{-0.25 cm}
\begin{figure}[ht]
\centerline{\includegraphics[width=.5\textwidth]{figures/bias_mapc.pdf}}
\caption{Differences in percent cropland estimates between the reference map and each of the four landcover maps. Rows indicate the landcover map being assessed (by subtraction from the reference map), while columns refer to resolution of aggregation. White indicates areas with no data where communal farmlands or plantation forests were removed.}\label{afoto}
\end{figure}

The spatial patterns of bias are most pronounced in the MODIS and GlobCover maps, with both substantially underestimating the amount of cropland in most regions of the country, particularly in the central part of the country where the bulk of the country's cropland is found (blue areas in Fig. 1). The mean of these differences, which is the landcover map's bias, are 21\% and 34\% at 1 km for MODIS and GlobCover, respectively (Fig. S1), meaning that each map underestimates cropland by that amount at this resolution. MODIS bias drops to 8\% at 50 km of aggregation, whereas GlobCover bias is still 24\% at 100 km (Fig. S1).

The SA-LC and GeoWiki datasets show much spatial variability and overall bias.
